\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-06 analysis/paper/paper.tex

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
functionality.

@@

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of stepwise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-9-1} \end{center}

@@

This report was generated on 2020-08-05 11:28:49 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] scales_1.1.1              forcats_0.5.0            
#>  [3] stringr_1.4.0             dplyr_1.0.0              
#>  [5] purrr_0.3.4               readr_1.3.1              
#>  [7] tidyr_1.0.2               ggplot2_3.3.2            
#>  [9] tidyverse_1.3.0           lubridate_1.7.8          
#> [11] visreg_2.6.1              tibble_3.0.3             
#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [15] flexsurv_1.1.1            survival_3.1-12          
#> [17] pracma_2.2.9              magrittr_1.5             
#> [19] kableExtra_1.1.0          glmnet_3.0-2             
#> [21] Matrix_1.2-18             data.table_1.12.8        
#> [23] cowplot_1.0.0             colorspace_1.4-1         
#> [25] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.4.1           
#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
#> [10] mvtnorm_1.1-0       xml2_1.3.2          codetools_0.2-16   
#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.6.1     
#> [16] broom_0.5.6         cluster_2.1.0       dbplyr_1.4.3       
#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.1         
#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
#> [28] tools_4.0.2         gtable_0.3.0        glue_1.4.1         
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     htmlwidgets_1.5.1   cmprsk_2.2-9       
#> [64] tidyselect_1.1.0    deSolve_1.28        R6_2.4.1           
#> [67] generics_0.0.2      Hmisc_4.4-0         multcomp_1.4-13    
#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
#> [73] haven_2.2.0         foreign_0.8-80      mgcv_1.8-31        
#> [76] nnet_7.3-14         mstate_0.2.12       modelr_0.1.6       
#> [79] crayon_1.3.4        rmarkdown_2.1       timereg_1.9.4      
#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
#> [91] viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
#> Head:     [d30ba8c] 2020-07-22: Merge pull request #23 from sahirbhatnagar/editMax
\end{CodeOutput}
\end{CodeChunk}

>>>>

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
functionality.

@@

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of stepwise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features. Competing Risks: whether or not an implementation for competing risks is present. Allows Non-PH: permits models for non-proportional hazards. Penalized Regression: allows for a penalty term when estimating hazards (e.g. lasso or ridge). Splines: permits a flexible fit on time through the use of splines. Parametric: implementation for parametric models. Semi-parametric: implementation for semi-parametric models. Interval/left censoring: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. Absolute Risk: computation for survival curves, cumulative incidence or cumulative hazard is available.}

@@

\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-9-1} \end{center}

@@

This report was generated on 2020-08-06 17:18:33 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.0              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.0.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.8          
#> [15] visreg_2.6.1              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.1-12          
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_3.0-2             
#> [25] Matrix_1.2-18             data.table_1.12.8        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [73] withr_2.2.0         pillar_1.4.6        haven_2.2.0        
#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
#> [79] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
#> [82] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.2         
#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
#> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    Jesse-Islam-master /Users/maxturgeon/Documents/git_repos/cbpaper
#> Head:     [32e6bf4] 2020-08-05: Merge pull request #25 from sahirbhatnagar/editMax
\end{CodeOutput}
\end{CodeChunk}
